msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-06-07 paper.tex

@@

\subsection*{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and
GTR~\citep{tavare1986some} nucleotide models
and
the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.

\textbf{TODO} Compare performance and features with
Seq-Gen~\citep{rambaut1997seq},
Dawg~\citep{cartwright2005dna} and Pyvolve~\citep{spielman2015pyvolve}

@@

\subsection*{Output}
The output of \msprime\ is a ``succinct tree sequence`` (or tree sequence,
for short) using the \tskit\ library. [Quick overview of the output formats
of other simulators, and the perf advantages of tree sequences.]

@@

>>>>

\subsection*{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and
GTR~\citep{tavare1986some} nucleotide models
and
the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}, and are run before every
release.

Comparing the performance of mutation generation methods is problematic.
Because there is no standard format for representing trees and
the accompanying mutations, all existing methods are focused on
writing out the resulting sequences. Methods are also not modular,
and so it is very difficult to disentangle the time required to
read in the tree file and output the sequences from the actual
time required to perform the simulation.

The mutation generation machinery in \msprime\ is extremely
efficient. For example, we simulated mutations under the JC69
model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim 280000$ mutations) in less that 1.5 seconds, including
the time required for file input and output. While a systematic
comparison of run-times is not feasible here, this is orders
of magnitude faster than methods such as
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}
and at least competitive with the recent method from
\cite{demaio2021phastsim}.

\textbf{TODO: would love to do a like-for-like comparison, but quite hard}

Another substantial advantage of our approach is that the simulated
mutations are stored directly in the output data format with
a defined API. Although
Pyvolve and Phastsim are implemented in Python, neither provide
programatic access to the simulated output, outputting instead
to a series of files which must subsequently be parsed.
See the Data Model section for more details and discussion.

@@

@@

\subsection*{Data model}
The output of \msprime\ is a ``succinct tree sequence`` (or tree sequence,
for short) using the \tskit\ library. [Quick overview of the output formats
of other simulators, and the perf advantages of tree sequences.]

[Review the data model in Phastsim~\cite{demaio2021phastsim} and
Usher~\citep{turakhia2021ultrafast}.]

@@

In terms of mutation simulation, future work will need to do things like
Dawg~\citep{cartwright2005dna} and to implement
codon models and indels~\citep{fletcher2009indelible}.
